# Practical: four

```{r load libraries and data, echo = FALSE ,include=FALSE}
library(tidyverse)
library(dplyr)
library(nycflights13)
data("flights")

```

## Question 1

```{r}
# This function displays columns and notes NA values

UtilsDataRSV::view_cols(flights)
```

## Question 2

```{r}


flight1 <- flights%>%
  filter(month == 1)%>%      # choosing all rows with a month value of 1
  group_by(carrier)%>%       # grouping by carrier to show carrier stats
  summarise(mean_distance = mean(distance),
            sd_distance = sd(distance))
flight1
```

## Question 3

The output above illustrates that carrier AS, HA, F9, YV and OO need to be investigated. The results shown below indicate that there are repeat values which is why the distance standard deviation for AS, HA, F9 and YV is zero.

```{r}

# From the output, the carriers with mean 0 and NA are AS, HA, F9, YV and OO


odd_carriers <- tibble(carrier = c("AS","HA", "F9", "OO", "YV"))


flights%>%
  filter(month == 1 )%>%
  semi_join(odd_carriers, join_by(carrier))%>%    # To leave desired carriers in tibble
  distinct(carrier, distance)                     # Shows the different values carriers have
```


The distance standard deviation of OO is NA and the output below shows that this is because there exists only one value during month 1. As a result the function `sd()` produces `NA`.

```{r}
#Showing how many entries of OO exist

flights%>%
  filter(carrier == "OO" & month == 1 )%>%
  select(carrier, month)
```

## Question 4


```{r}

flights %>%
  group_by(month, carrier) %>%   # grouping by two variables so they appear in final output
  summarise(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE)
  ) %>%
  pivot_wider(names_from = carrier, values_from = avg_dep_delay)
```

## Question 5

```{r}
# tibble which meets conditions

flight_condition<-  flights%>%
  filter(dep_delay > 0 & arr_delay <= 0)

total_flights <- nrow(flights)           # number of total flights
met_condition <- nrow(flight_condition)  # number of flights which met condition

prop_of_flights <- met_condition / total_flights 
print(prop_of_flights)
```

## Question 6: Part 1 and 2

```{r}


multi_routes <- flights %>%            
  group_by(origin, dest) %>%            # grouping allowing for count in the next line  
  filter(n_distinct(carrier) > 1) %>%   # filtering out non-repeat routes 
  ungroup()

avg_arr_delay <<- multi_routes %>%
  group_by(origin, dest, carrier) %>%   
  summarise(
    arr_del_avg = mean(arr_delay, na.rm = TRUE),
    .groups = "drop"
  )

avg_arr_delay

```

## Question 6: Part 3

```{r}

# filter is used to keep the carrier with lowest delay time per route

avg_arr_delay %>% 
  group_by(origin, dest) %>%            
  filter(arr_del_avg == min(arr_del_avg, na.rm = TRUE)) %>%  
  ungroup()

```

## Question 6: Part 4


```{r}

avg_arr_delay %>%
  group_by(origin, dest) %>%
  summarise(
    best_delay  = min(arr_del_avg, na.rm = TRUE), # smallest delay by route
    worst_delay = max(arr_del_avg, na.rm = TRUE), # largest delay by route
    diff_del    = worst_delay - best_delay,        
    .groups = "drop"
  )%>%
  filter(diff_del == max(diff_del, na.rm =TRUE)) # retain only the largest delay


```

## Question 6: Part 5

From the previous output it is evident that the JFK-ATL route has the largest difference so it is investigated below. 

```{r}

# Observing JFK ATL flights with their carriers


avg_arr_delay %>%
  filter(origin =="JFK", dest =="ATL")


```

As a result, carriers 9E, DL and EV are assessed below. The output shows that there was a flight a flight by carrier EV which was delayed more than normal. The other carriers with more flights have less extreme arrival delay times. Therefore, the reason for this large number is that there was an outlier flight by carrier EV.

```{r}

# Filter flights for JFK → ATL and carriers 9E, DL, EV 

flights %>%
  filter(origin == "JFK", dest == "ATL", carrier %in% c("9E", "DL", "EV"))%>%
  group_by(carrier) %>%
  summarise(
    num_flights = n(),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    median_arr_delay = median(arr_delay, na.rm = TRUE),
    max_arr_delay = max(arr_delay, na.rm = TRUE),
    min_arr_delay = min(arr_delay, na.rm = TRUE),
    .groups = "drop"
  )           # obtaining flight data for assessment 
```


## Question 7

The output shows that height and glucose have missing values. Furthermore, disease status and gender have incorrect entries.

```{r}

# Storing into data frame

error_data <- structure(list(
  id = c("id_1", "id_2", "id_3", "id_4", "id_5", "id_6", "id_7", "id_8", "id_9", "id_10",
         "id_11", "id_12", "id_13", "id_14", "id_15", "id_16", "id_17", "id_18", "id_19", "id_20",
         "id_21", "id_22", "id_23", "id_24", "id_25", "id_26", "id_27", "id_28", "id_29", "id_30",
         "id_31", "id_32", "id_33", "id_34", "id_35", "id_36", "id_37", "id_38", "id_39", "id_40",
         "id_41", "id_42", "id_43", "id_44", "id_45", "id_46", "id_47", "id_48", "id_49", "id_50"),
  age = c(50L, 34L, 70L, 33L, 22L, 61L, 69L, 73L, 62L, 56L, 71L, 33L, 73L, 44L, 45L, 46L, 24L,
          70L, 46L, 76L, 47L, 76L, 28L, 48L, 54L, 27L, 45L, 26L, 61L, 28L, 38L, 55L, 33L, 36L,
          62L, 58L, 72L, 31L, 34L, 51L, 61L, 64L, 26L, 28L, 60L, 29L, 42L, 46L, 79L, 72L),
  gender = c("male", "male", "male", "female", "female", "male", "female", "male", "male", "female",
             "female", "male", "male", "female", "male", "male", "male", "male", "female", "male",
             "male", "male", "male", "female", "femal", "male", "female", "female", "female", "female",
             "male", "female", "female", "female", "male", "male", "female", "male", "female", "female",
             "male", "female", "female", "male", "male", "female", "male", "male", "male", "female"),
  height = c(174.4, 197.7, 174.1, 194.5, NA, 180.4, 170.5, 157.4, 196.8, 165.1,
             153, 197.4, 186, 157.1, 177.5, 197.7, 179.3, 170.2, 182.4, NA,
             165.4, 161, 168.5, 199.2, 157.7, 154.6, 157.1, 184.5, 181, 194.6,
             183.6, 186.9, 176.1, 183, 191.1, 189.3, 199, 172, 165.6, 170.5,
             150.5, 159.2, 192.1, 161.6, 162, 153.8, 162.3, 186.6, 192.4, 174.9),
  weight = c(69.4, 62.3, 55.6, 69.5, 78.6, 60.8, 72.2, 60.9, 75.1, 67.7,
             82.5, 68.7, 67.8, 76.7, 87, 61.1, 70.6, 63.3, 81.5, 59.2, 93.2,
             87.3, 83.4, 80.9, 68.6, 76.5, 93.7, 79.1, 92, 65.6, 85.4, 63.3,
             79.7, 74.1, 63.3, 78.2, 95.7, 95.1, 63.7, 66.1, 99.3, 81, 96.9,
             73.3, 70.3, 83, 57.6, 78.6, 61.9, 98.1),
  disease_status = c("diseased", "healthy", "healthy", "healthy", "healthy", "healthy", "diseased",
                     "healthy", "diseased", "Healthy", "diseased", "healthy", "diseased", "healthy",
                     "diseased", "healthy", "healthy", "healthy", "healthy", "healthy", "healthy",
                     "diseased", "healthy", "diseased", "healthy", "healthy", "healthy", "healthy",
                     "diseased", "diseased", "healthy", "healthy", "healthy", "diseased", "diseased",
                     "diseased", "healthy", "diseased", "healthy", "healthy", "healthy", "healthy",
                     "healthy", "diseased", "diseased", "diseased", "healthy", "healthy", "diseased",
                     "diseased"),
  glucose = c(96, 78, 101, 119, 103, 91, 86, NA, 77, 80, 115, 85, 88, 109, NA, 71, 90, 94, 91,
              87, 113, 93, 97, 118, 109, 80, 85, 119, 99, 108, 89, 108, 97, 116, 79, 84, 75, 81,
              119, NA, 106, 109, 75, 82, 84, 75, 76, 120, 119, 77)
), row.names = c(NA, -50L), class = c("tbl_df", "tbl", "data.frame"))

# This viewing option shows details of the columns

UtilsDataRSV::view_cols(error_data)
```
